\paperTitle

### 2014-05-06 doc/icml2014/assumption-proof.tex
Complete tensor proofs

% Define up front that we will focus on h, x_1, x_2, x_3
In step 1 of \LearnMarginals, we used the bottleneck property of a hidden
  variable $h_i$ to learn conditional moments $\mOpp{v}{i}$ for every
  view $x_v \in \sV_{h_i}$ using \TensorFactorize. 
In order to do so, we require that \assumptionref{full-rank} holds, i.e.
\begin{assumption*}
  Given a bottleneck $h_1$ with views $x_1, x_2, x_3$, the conditional
  moments $\mOpp{1}{1}, \mOpp{2}{1}, \mOpp{3}{1}$ have full column rank
  $k$, and $\pi\oft 1 \succ 0$.
\end{assumption*}

To interpret this assumption better, we would like find an equivalent
condition on the model parameters $Z_\sC$ that imply the above condition
on $\mOpp{v}{i}$. In this section, we will show that
\assumptionref{full-rank} is actually implied by the following condition
(\assumptionref{full-rank-plus});
\begin{assumption*}
For every clique $\sC \in \sG$ (including ones involving observed
variables), every mode-unfolding of the marginals $Z_\sC$ has full
column rank.
\end{assumption*}

@@

\begin{figure}
  \label{fig:message-proof}
  \caption{A recursive construction of the conditional moments $\mOpp{v}{i}$.}
\end{figure}

@@

Going forward, we will construct $\mYpp{v}{1}$ recursively.
Without loss of generality, let $h_1, h_2, \cdots, h_t$ be a topological
  ordering rooted at $h_1$,
and let $\Pa(h)$ be the parents of a hidden variable $h$ in
  this topological ordering.
Then $\mYpp{t}{1}$ can be expressed as,
\begin{align*}
  \mYpp{t}{1} &\eqdef \Pr( h_t \given h_1 )  \\
  &= \sum_{\vh \in H_{\Pa(h_t)}} \Pr( \Pa(h_t) \given h_1 ) \Pr( h_t \given \Pa(h_t) ) \\
  &= \mYpp{ \Pa(h_t) }{1} \times_{\Pa(h_t)} \mYpp{t}{\Pa(h_t)},
\end{align*}
where $A \times_{C} B$ refers to summation along the indices $C$. We
refer to the operation $A \times_C B$ as tensor multiplication.
\appendixref{tensor-multiplication} prooves some key properties of this
operation. 
Note that $\mYpp{t}{\Pa(h_t)}$ is the conditional probability $\Pr(h_t
 | \Pa(h_t))$, which can be easily computed from the hidden marginals
 $Z_\sC$ that contain $h_t$. 
We will describe this procedure in further detail soon.

Now, let's further expand $\mYpp{ \Pa(h_t) }{1}$ in terms of {\em its
  parents}, recursively doing so until we reach $h_1$.
Let $C$ be an intermediate set of hidden variables. Let $h_c \neq h_1$
  be some hidden variable in $C$ which we wish to eliminate. 
Let $C' = C \union \Pa(h_c) \setminus \{h_c\}$ and $\del C = C'
  \setminus (C \union \{h_1\})$, the variables in the interface between $C$ and $C'$.
Then,
\begin{align*}
  \mYpp{C}{1} &\eqdef \Pr( h_C \given h_1 )  \\
  &= \sum_{\vh \in H_{\del C}} \Pr( C' \given h_1 ) \Pr( h_c \given \Pa(h_c) ) \\
  &= \mYpp{ C' }{ 1 } \times_{\del C} \mYpp{c}{\Pa(h_c)},
\end{align*}
Finally, for the base case, $\mYpp{1}{1} = \ones$.
This procedure is algorithmically described in \algorithmref{Y}.

@@

Note that in the process of the algorithm, $\mYpp{t}{1}$ is expressed
entirely as tensor multiplications with terms of the form
$\mYpp{c}{\Pa(h_c)}$ for some hidden variable $h_c$.  Let's finally
derive a concrete expression for $\mYpp{c}{\Pa(h_c)}$ from the
parameters, the hidden marginals $Z_\sC$.

Without loss of generality, let $\sC \eqdef \{c\} \union P$ be a clique
  in the graph $\sG$.
If instead $\sC \subseteq \sC_1 \union \sC_2$ for some $\sC_1, \sC_2 \in
  \sG$, we can simply construct $Z_{\sC}$ from the union of $\sC_1$ and
  $\sC_2$,
\begin{align*}
  Z_\sC &= Z_{\sC_1} \times_{\del \sC} Z_{\sC_2},
\end{align*}
where $\del \sC = (\sC_1 \union \sC_2) \setminus \sC$.

Proceeding,
\begin{align*}
\mYpp{c}{\Pa(h_c)} 
  &\eqdef \Pr(h_c \given \Pa(h_c)) \\
  &= \frac{\Pr(h_c, \Pa(h_c))}{\Pr(\Pa(h_c))} \\
  &= Z_\sC \diag{Z_\sC(\ones, \cdot, \ldots, \cdot)}^{-1}.
\end{align*}

\begin{lemma}[Sufficiency of \assumptionref{full-rank-plus}]
  \label{lem:full-rank-suff}
  Given that \assumptionref{full-rank-plus} holds, then for any hidden
  variable $h_1$ and observed variable $x_v$, $\mPi{1} \succ 0$ and
  $\mOpp{v}{i}$ has full column rank.
\end{lemma}
\begin{proof}
  The important statement that needs to be made is how $\sigma_k(Y)$
  relates to its parents.
\end{proof}

>>>>

% Define up front that we will focus on h, x_1, x_2, x_3
In step 1 of \LearnMarginals, we used the bottleneck property of a hidden
  variable $h_i$ to learn conditional moments $\mOpp{v}{i} \eqdef
  \Pr(x_v | h_i)$ for every view $x_v \in \sV_{h_i}$ using
  \TensorFactorize. 
In order to do so, we require that \assumptionref{full-rank} holds, i.e.
\begin{assumption*}
  Given a bottleneck $h_1$ with views $x_1, x_2, x_3$, the conditional
  moments $\mOpp{1}{1}, \mOpp{2}{1}, \mOpp{3}{1}$ have full column rank
  $k$, and $\pi\oft 1 \succ 0$.
\end{assumption*}

To interpret this assumption better, we would like find an equivalent
condition on the model parameters $Z_\sC$ that imply the above condition
on $\mOpp{v}{i}$. In this section, we will show that
\assumptionref{full-rank} is actually implied by the following condition
(\assumptionref{full-rank-plus});
\begin{assumption*}
For every clique $\sC \in \sG$ (including ones involving observed
  variables), for every {\em conditioning} of the marginals, every
  unfolding has full rank. 
\end{assumption*}

@@

\begin{figure}
  \caption{A recursive construction of the conditional moments $\mOpp{v}{i}$.}
  \label{fig:message-proof}
\end{figure}

@@

Going forward, we will construct $\mYpp{v}{1}$ recursively.
Without loss of generality, let $h_1, h_2, \cdots, h_t$ be a topological
  ordering rooted at $h_1$,
and let $\Pa(h)$ be the parents of a hidden variable $h$ in
  this topological ordering.
Then $\mYpp{t}{1}$ can be expressed as,
\begin{align*}
  \mYpp{t}{1} &\eqdef \Pr( h_t \given h_1 )  \\
  &= \sum_{\vh \in H_{\Pa(h_t)}} \Pr( \Pa(h_t) \given h_1 ) \Pr( h_t \given \Pa(h_t) ) \\
  &= \mYpp{ \Pa(h_t) }{1} \times_{\Pa(h_t)} \mYpp{t}{\Pa(h_t)},
\end{align*}
where $A \times_{C} B$ refers to summation along the indices $C$. We
refer to the operation $A \times_C B$ as tensor multiplication.
\appendixref{tensor-multiplication} prooves some key properties of this
operation. 
Note that $\mYpp{t}{\Pa(h_t)}$ is the conditional probability $\Pr(h_t
 | \Pa(h_t))$, which is assumed to be full rank, and can be easily
 computed from the hidden marginals $Z_\sC$ that contain $h_t$.

Now, let's further expand $\mYpp{ \Pa(h_t) }{1}$ in terms of {\em its
  parents}, recursively doing so until we reach $h_1$.
Let $C$ be an intermediate set of hidden variables. Let $h_c \neq h_1$
  be some hidden variable in $C$ which we wish to eliminate. 
Let $C' = C \union \Pa(h_c) \setminus \{h_c\}$, the new set of hidden
  variables containing the parents of $h_c$ instead of $h_c$ and $\del
  C = C' \setminus (C \union \{h_1\})$, the variables in the interface
  between $C$ and $C'$.
Then,
\begin{align*}
  \mYpp{C}{1} &\eqdef \Pr( h_C \given h_1 )  \\
  &= \sum_{\vh \in H_{\del C}} \Pr( C' \given h_1 ) \Pr( h_c \given \Pa(h_c) ) \\
  &= \mYpp{ C' }{ 1 } \times_{\del C} \mYpp{c}{\Pa(h_c)},
\end{align*}
Finally, for the base case, $\mYpp{1}{1} = \ones$.
This procedure is algorithmically described in \algorithmref{Y}.

@@

Note that in the process of the algorithm, $\mYpp{t}{1}$ is expressed
entirely as tensor multiplications with terms of the form
$\mYpp{c}{\Pa(h_c)}$ for some hidden variable $h_c$. 
$\mYpp{c}{\Pa(h_c)}$ can be easily computed from the clique marginals
$Z_\sC$.

Without loss of generality, let $\sC \eqdef \{c\} \union \Pa(h_c)$ be a clique
  in the graph $\sG$.
If instead $\sC \subseteq \sC_1 \union \sC_2$ for some $\sC_1, \sC_2 \in
  \sG$, we can simply construct $Z_{\sC}$ from the union of $\sC_1$ and
  $\sC_2$,
\begin{align*}
  Z_\sC &= Z_{\sC_1} \times_{\del \sC} Z_{\sC_2},
\end{align*}
where $\del \sC = (\sC_1 \union \sC_2) \setminus \sC$.

Proceeding,
\begin{align*}
\mYpp{c}{\Pa(h_c)} 
  &\eqdef \Pr(h_c \given \Pa(h_c)) \\
  &= \frac{\Pr(h_c, \Pa(h_c))}{\Pr(\Pa(h_c))} \\
  &= Z_\sC \times_{\Pa(h_c)} \diag{Z_\sC(\ones, \cdot, \ldots, \cdot)}^{-1},
\end{align*}
where $\diag(\sP)$ is a ``diagonal'' tensor constructed by taking the
tensor product with $I^{\otimes p}$,
\begin{align*}
  \diag(\sP)[\vi,\vj] &\eqdef \sP[\vi] \delta[\vi,\vj].
\end{align*}

\paragraph{Sample complexity results}

From the algorithm described above, blah.

\begin{lemma}[Sufficiency of \assumptionref{full-rank-plus}]
  \label{lem:full-rank-suff}
  Given that \assumptionref{full-rank-plus} holds, then for any hidden
  variable $h_1$ and observed variable $x_v$, $\mPi{1} \succ 0$ and
  $\mOpp{v}{i}$ has full column rank.
\end{lemma}
\begin{proof}

  - $Y$
  - Use theorem of tensor multiplication.
\end{proof}

The sample complexity of estimating $\mOpp{v}{i}$ depends mainly on
$\sigma_{\min}(\mOpp{v}{i})$.

\begin{lemma}[Singular values of $\mOpp{v}{1}$]
  \label{lem:mopp-singular-values}

Let $h_t$ be the unique parent of $x_v$, as above.
Let $h_1 \succ h_2 \cdots \succ h_t$ be a topological ordering of
  variables according to the graph $\sG$.
  Then, the smallest singular value of $\mOpp{v}{1}$ is at least the product of the smallest singular values of $\mYpp{c}{\Pa(h_c)}$.
\begin{align*}
  \sigma_{\min}(\sP) &\ge \prod_{c \in [t]} \sigma_{\min}(\mYpp{c}{\Pa(h_c)}),
\end{align*}
where $h_c$

\end{lemma}
\begin{proof}
  - $Y$
  - Use theorem of tensor multiplication.
\end{proof}

\begin{lemma}[Sample complexity for $\mOpp{v}{1}$]
  \label{lem:sample-complexity}
  The sample complexity to estimate $\mOpp{v}{i}$ via the tensor power method \citep{anandkumar13tensor} is 
\end{lemma}
\begin{proof}
  The result follows directly from Theorem x of \citet{anandkumar13tensor}.
\end{proof}

### 2014-05-06 doc/icml2014/tensor-multiplication.tex
Complete tensor proofs

\paragraph{Unfoldings}
For a tensor $\sP$, as defined above, 
  given an index set $I \subset [m]$ we
  define an unfolding of a tensor $\sP$ along $I$,
 $\sP\munf{I}$, to be a matrix of dimensions $(\prod_{i \in I} d_i)
 \times (\prod_{j \not\in I} d_j)$. $\sP\munf{I}$ has elements
 $\sP\munf{I}[\vi;\vj] = \sP[\vi\cdot\vj]_{\pi}$, where $\pi$ is an
 appropriate permutation to align the indices of $\vi$ and $\vj$ with
 the ordering of $I$. 
For example, if $\sP$ were a $4$-th order tensor and $I = \{1,3\}$,
  $\sP\munf{13}[i_1,i_2; j_1,j_2] = \sP[i_1,j_1,i_2,j_2]$.
For notational convenience, we will omit the permutation $\pi$ subsequently.

@@

\begin{lemma}[Tensor Multiplication and Singular Values]
  \label{lem:tensor-multiplication}
Let $\sP, \sQ$ be $m$-th order and $n$-th order tensors respectively, $M$ be an
  index mapping and $\sR = \sP \times_M \sQ$. 
Let $I \subset [m - k + n - k]$ be any arbitrary subset of the index set
  of $\sR$,
and let $I_P$, $I_Q$ be the partitioning of $I$ into subsets contained
  in $\sP$ and $\sQ$ respectively.
Then,
\begin{align*}
\sigma_{\max}(\sR\munf{I}) &\le \sigma_{\max}(\sP\munf{I_P}) \sigma_{\max}(\sQ\munf{I_P}) \\
\sigma_{\min}(\sR\munf{I}) &\ge \sigma_{\min}(\sP\munf{I_Q}) \sigma_{\min}(\sQ\munf{I_Q}).
\end{align*}
\end{lemma}

@@

Consider an $(I)$-unfolding of $\sR$,
  \begin{align*}
    \sR\munf{I}[\vi;\vj] 
       &= \sum_{\vk \in D_\ell} \sP[\vk\cdot \vi_P \cdot \vj_P] \sQ[\vk\cdot \vi_Q \cdot \vj_Q] \\
       &= \sum_{\vk \in D_\ell} \sP\munf{I_P}[\vi_P; \vj_P \cdot \vk] \sQ\munf{I_Q}[\vj_Q; \vj_Q \cdot \vk].
  \end{align*}
Note that the indices that were summed out belong in the complement of $I_P$ and $I_Q$ respectively.

The key idea is to represent the above expression as matrix multiplication.
To do so, we can temporarily ``inflate'' the unfoldings $\sP\munf{I_P}$
  and $\sQ\munf{I_Q}$ by taking the outer product of $\sP\munf{I_P}$ with
  $|I_Q|$ identity matrices, and vice-versa. Define 
\begin{align*}
  \sP' &\eqdef \sP \otimes I_{d'_1} \otimes \cdots \otimes I_{d'_{|I_Q|}} \\
  \sQ' &\eqdef \sQ \otimes I_{d'_1} \otimes \cdots \otimes I_{d'_{|I_P|}}.
\end{align*}

Now, $\sP'$ is a $(m + 2 |I_Q|)$-th order tensor and similarly $\sQ'$
  is a $(n + 2 |I_P|)$-th order tensor. 
For every identity matrix we tensor $\sP$ and $\sQ$ by we include one of
  its indices to include in the new set $I'_P$ and $I'_Q$ respectively.
  Note that $\sP\munf{I'_P}$ is a $(I_P + I_Q)\times (m - |I_P| + |I_Q| + \ell)$ matrix, and 
$\sP\munf{I'_Q}$ is $(I_P + I_Q) \times (n - |I_Q| + |I_P| + \ell)$
Then,
\begin{align*}
    \sR\munf{I}[\vi;\vj] 
    &= \sum_{\be_P, \be_Q \vk}
    {\sP'}\munf{I'_P}[\vi_P\be'_Q; \vj_P \be_Q \vk]  
    {\sQ'}\munf{I'_Q}[\vi_Q\be'_P; \vj_Q \be_P \vk] \\
    \sR\munf{I} &= {\sP'}\munf{I'_P} {\sQ'}\munf{I'_Q}.
\end{align*}

>>>>

\paragraph{Unfoldings}
For a tensor $\sP$, as defined above, 
  given an index set $I \subset [m]$ we
  define an unfolding of a tensor $\sP$ along $I$,
 $\sP\munf{I}$, to be a matrix of dimensions $(\prod_{i \in I} d_i)
 \times (\prod_{j \not\in I} d_j)$. $\sP\munf{I}$ has elements
 $\sP\munf{I}[\vi;\vj] = \sP[\vi\cdot\vj]_{\pi}$, where $\pi$ is an
 appropriate permutation to align the indices of $\vi$ and $\vj$ with
 the ordering of $I$. 
For example, if $\sP$ were a $4$-th order tensor and $I = \{1,3\}$,
  $\sP\munf{13}[i_1,i_2; j_1,j_2] = \sP[i_1,j_1,i_2,j_2]$.
For notational convenience, we will omit the permutation $\pi$ subsequently.
Let $D_I = \bigtimes_{i \in I} [d_i]$ and  $\bar I \eqdef [m] \setminus
  I$ to represent the complement of indices. Note that $\sP\munf{I}
  = (\sP\munf{\bar I})^\top$.

@@

\begin{theorem}[Tensor Multiplication and Singular Values]
  \label{thm:tensor-multiplication}
Let $\sP, \sQ$ be $m$-th order and $n$-th order tensors respectively, $M$ be an
  index mapping and $\sR = \sP \times_M \sQ$. 
Let $I \subset [m - k + n - k]$ be any arbitrary subset of the index set
  of $\sR$,
and let $I_P$, $I_Q$ be the partitioning of $I$ into subsets contained
  in $\sP$ and $\sQ$ respectively.
Then,
\begin{align*}
\sigma_{\max}(\sR\munf{I}) &\le \sigma_{\max}(\sP\munf{I_P}) \sigma_{\max}(\sQ\munf{I_P}) \\
\sigma_{\min}(\sR\munf{I}) &\ge \sigma_{\min}(\sP\munf{I_Q}) \sigma_{\min}(\sQ\munf{I_Q}).
\end{align*}
\end{theorem}

@@

Consider an $(I)$-unfolding of $\sR$,
  \begin{align*}
    \sR\munf{I}[\vi;\vj] 
       &= \sum_{\vk \in D_\ell} \sP[\vk\cdot \vi_P \cdot \vj_P] \sQ[\vk\cdot \vi_Q \cdot \vj_Q] \\
       &= \sum_{\vk \in D_\ell} \sP\munf{I_P}[\vi_P; \vj_P \cdot \vk] \sQ\munf{\bar I_Q}[\vj_Q ; \vi_Q \cdot \vk].
  \end{align*}
Note that the indices that were summed out belong in the complement of
$I_P$ and $I_Q$ respectively.

The key idea is to represent the above expression as matrix multiplication.
To do so, we can temporarily ``inflate'' the unfoldings $\sP\munf{I_P}$
  and $\sQ\munf{I_Q}$ by taking the outer product of $\sP\munf{I_P}$ with
  $|I_Q|$ identity matrices, and vice-versa. Define 
\begin{align*}
  \sP' &\eqdef \sP \otimes I_{d'_1} \otimes \cdots \otimes I_{d'_{|I_Q|}} \\
  \sQ' &\eqdef \sQ \otimes I_{d'_1} \otimes \cdots \otimes I_{d'_{|\bar I_P|}}.
\end{align*}

Now, $\sP'$ is a $(m + 2 |I_Q|)$-th order tensor and similarly $\sQ'$
  is a $(n + 2 |\bar I_P|)$-th order tensor. 
For every identity matrix we tensor $\sP$ and $\sQ$ by, we include one of
  its indices in the new set $I'_P$ and $I'_Q$ respectively.
Note that ${\sP'}\munf{I'_P}$ is a $(I_P + I_Q)\times (|\bar I_P| + \ell
  + |I_Q|)$ matrix, and ${\sQ'}\munf{\bar I'_Q}$ is $(|\bar I_Q| + |\bar
  I_P|) \times (|\bar I_Q| + \ell + |\bar I_P|)$ matrix.
Then,
\begin{align*}
    \sR\munf{I}[\vi;\vj] 
    &= \sum_{\be_P, \be_Q, \vk}
    {\sP'}\munf{I'_P}[\vi_P \cdot \be'_Q; \vj_P \cdot \vk \cdot \be_Q ]  \\ 
    &\quad \hphantom{\sum_{\be_P, \be_Q, \vk}}
    {\sQ'}\munf{I'_Q}[\vi_Q \cdot \be'_P; \vj_Q \cdot \vk \cdot \be_P ] \\
    \sR\munf{I} &= {\sP'}\munf{I'_P} ({\sQ'}\munf{\bar I'_Q})^\top \\
                &= {\sP'}\munf{I'_P} {\sQ'}\munf{I'_Q}.
\end{align*}
